
```{r include = FALSE}
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```
# External usage {-}

## How to adopt this course {-}

Not all data used within the practicals is distributed on GitHub due to size limitations. Each practical describe how to access the datasets used within.

There are two main options to adopt this course:

1. Adopt the course in its entirety:
  + Install R and RStudio
  + Clone the project from GitHub to RStudio (in RStudio, File > New Project > Version Control > Git)
  + Install the  `renv` package and install all packages within the project using `renv::restore()`. On the project's GitHub the `renv.lock` file contains all of the packages (and versions) used within it.  
  + If you wish to build the book at any point, practicals (RMarkdown files) without data can be removed by commenting out the file (with a #) in the `_bookdown.yml` file.
  
For more details on how to use Git, GitHub and `renv` see the practical [Git, GitHub and RMarkdown]

2. Adopt a single practical by downloading the `.rmd` file, packages loaded within the `.rmd` file and associated data from the links listed within it.  

## How to contribute {-}

We welcome contributions to the material either directly, for any small issues or errors (e.g. typos) or through a GitHub issue for larger issues (e.g. further explanation requests, bugs in the code)

**For smaller issues** click the <i class ="fa fa-edit fa-lg"></i> button in the top tool bar: 

* Create a fork of the repository and clone it to your local computer
* Create a local branch 
* Make changes to that branch (commit them)
* Push the changes to your GitHub fork
* Once you have made then [open a pull request](https://github.com/andrewmaclachlan/CASA0005repo/pulls) for it to be considered. 

**For larger issues** [open a issue on the project's GitHub issue page](https://github.com/andrewmaclachlan/CASA0005repo/issues). 

If raising an issue please provide sufficient detail that it can be reproduced with a minimal worked example (where possible). Be sure to include the following:

* A short explanation of the bug and it's environment (e.g. operating system).
* A step by step guide to reproduce it.

**For clarification on concepts** please either [open an issue on the project's GitHub issue page](https://github.com/andrewmaclachlan/CASA0005repo/issues) or submit the anonymous feedback form at the end of each practical. 

Alternatively, reach out to us on Twitter:

* [Andy MacLachlan](https://twitter.com/andymaclachlan)
* [Adam Dennett](https://twitter.com/adam_dennett)

## License {-}

If you use this material for teaching, research or anything else please let me (Andy) know via [Twitter](https://twitter.com/andymaclachlan) or email --- a [dot] maclachlan [at] ucl [dot] ac [dot] uk).

This practical book is licensed under a [Creative Commons Attribution-ShareAlike 4.0 International (CC BY-SA 4.0) License](https://creativecommons.org/licenses/by-sa/4.0/).
<center>
<i class="fab fa-creative-commons fa-2x"></i><i class="fab fa-creative-commons-by fa-2x"></i><i class="fab fa-creative-commons-sa fa-2x"></i>
</center>

You are free to: 

* **Share** — copy and redistribute the material in any medium or format

* **Adapt** — remix, transform, and build upon the material
for any purpose, even commercially.

However, you  give appropriate credit, provide a link to the license, and indicate if changes were made. If you remix, transform, or build upon the material, you must distribute your contributions under the same license as the original.

But, you do not have to comply with the license for elements of the material in the public domain or where your use is permitted by an applicable exception or limitation.

## Updates {-}

This is the third iteration of the practical book, representing the third year it has been hosted in this format.

Previous versions can be found at the following links.

2020-2021 academic year (on which the JOSE paper is based): https://andrewmaclachlan.github.io/CASA0005repo_20202021/

2019-2020 academic year: https://andrewmaclachlan.github.io/CASA0005repo_20192020/

A summary of changes is provided on the current README.md file: https://github.com/andrewmaclachlan/CASA0005repo

## Packages {-}

This bookdown document was built with:

```{r message = FALSE, warning = FALSE, echo = FALSE}
library(tidyverse)
library(devtools)
library(DT)

# needed because new_session is set to true in _bookdown.yml
all_pkgs <- renv::dependencies(path = "DESCRIPTION") %>% 
  pull(Package) 

session <- devtools::session_info(pkgs = all_pkgs)

session$platform
```

And makes use of the following packages throughout. Thanks to [STAT 545](https://stat545.com/index.html#other-contributors) for this code.

```{r message = FALSE, warning = FALSE, echo = FALSE}
library(dplyr)
session$packages %>% 
  dplyr::select(package, loadedversion, date, source) %>% 
  DT::datatable(rownames = FALSE,
                class = 'cell-border stripe',
                filter = list(position = 'top'),
                options = list(pageLength = nrow(session$packages), 
                           autoWidth = TRUE,
                           bInfo = FALSE,
                           paging = FALSE))
```
