# Interactive maps {-}

::: {.infobox .tip data-latex="{note}"}
This uses the same data from the mapping practical - airbnbs and hotels (OSM) in London. Run all of that code before starting this!
:::

```{r message=FALSE, echo=FALSE}
library(sf)
library(tidyverse)

# OSM data

OSM <- st_read("prac5_data/greater-london-latest-free.shp/gis_osm_pois_free_1.shp")%>%
  st_transform(., 27700) %>%
  #select hotels only
  filter(fclass == 'hotel')

# Airbnb data
Airbnb <- read_csv("prac5_data/listings.csv") %>%
  # longitude is considered x value here, latitude is y
  st_as_sf(., coords = c("longitude", "latitude"), 
                   crs = 4326) %>%
    st_transform(., 27700)%>%
    #select entire places that are available all year
    filter(room_type == 'Entire home/apt' & availability_365 =='365')


#London Borough data is already in 277000
Londonborough <- st_read("Prac1_data/statistical-gis-boundaries-london/ESRI/London_Borough_Excluding_MHW.shp")%>%
  st_transform(., 27700)

Accomodation_contained <- Londonborough%>%
  mutate(hotels_n = lengths(st_contains(., OSM)))%>%
  mutate(airbnbs_n = lengths(st_contains(., Airbnbs)))

library(classInt)

# Get Jenks breaks for 5 classes
breaks <- Accomodation_contained%>%
  st_drop_geometry()%>%
  #need a numeric vector not a dataframe or tibble
  pull(airbnbs_n) %>%             
  classIntervals(., n = 5, style = "jenks")
```

To make our map interactive we change `tmap_mode()` from plot to view..

```{r message=FALSE, warning=FALSE, cache=FALSE}
tmap_mode("view")

tm_shape(Accomodation_contained) + 
  tm_polygons("hotels_n") 
```

## Advanced interactive map

But let's take it a bit further so we can select our layers on an interactive map..Let's set a few more things up:

* Our CRS needs to be in WGS84

```{r}
Accomodation_containedWGS84<- Accomodation_contained %>%
  st_transform(., 4326)

```

Similar to our static maps we must:

* First define the shape object 
* Then load the map layer (e.g. polygons)

Here, we also set a group which links the polygon both the legend and button selection options.

```{r prac5leaf, message=FALSE, warning=FALSE, cache=FALSE, eval=TRUE}

tmap_mode("view")

# Build the map

  # Airbnb layer
map <-  tm_shape(Accomodation_containedWGS84) +
        tm_polygons(fill="airbnbs_n",
                    fill.scale = list(tm_scale_intervals(values = "brewer.blues", breaks=breaks$brks)),
                    fill_alpha=0.5,
                    col="white",
                    lwd=2,
                    lty="dashed",
                    # select columns for popup
                    popup.vars=c("airbnbs_n", "NAME"),
                    group="Airbnb") + 
        
  # hotel layer  
        tm_polygons(fill="hotels_n",
                    fill.scale = list(tm_scale_intervals(values = "brewer.blues", breaks=breaks$brks)),
                    fill_alpha=0.5,
                    col="white",
                    lwd=2,
                    lty="dashed",
                    popup.vars=c("hotels_n", "NAME"),
                    group="Hotels")+
    # basemap
          tm_basemap(server = c("OpenStreetMap", "Thunderforest.Landscape"))
        

map
```

To see other basemap options (there are loads!) have a look here at [leaflet extras](http://leaflet-extras.github.io/leaflet-providers/preview/)
